\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-13 tex/paper.tex
minor edits

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly without
use Kantorovich-Rubinstein duality and the critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``Unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MEWE by taking expectations with the
variational program $q(\bfx | \bf\theta)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the
critic $d$ allows us to work in high dimensions without summary statistics and
optimize $\bfpsi$ with stochastic gradient descent.
\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL. We also don't specify a prior b/c it's like implicit empirical bayes. They also don't specify a recognition model to latents. }

@@

This work has many connections to work on variational inference, in which the
goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
is close to the true posterior $p(\bfz, \bftheta |x)$. There have been efforts
to extend variational inference to intractable likelihoods; however, they
require restrictive assumptions.  In \citep{tran2017variational}, the authors
consider Variational Bayes with an Intractable Likelihood (VBIL). In that
approach "the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
$\epsilon > 0$,   thus this method inherits the drawbacks of the ABC-likelihood
including the choice of summary statistics and the inefficiency in evaluating
the ABC likelihood for high-dimensional data and small $\epsilon$.

@@

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.

>>>>

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \citep{2017arXiv170208235H}  are a recent extension to GANs that
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``Unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MEWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and
optimize $\bfpsi$ with stochastic gradient descent.
\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL. We also don't specify a prior b/c it's like implicit empirical bayes. They also don't specify a recognition model to latents. }

@@

This work has also many connections to work on variational inference, in which the
goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
to extend variational inference to intractable likelihoods; however, they
require restrictive assumptions.  In \citep{tran2017variational}, the authors
consider Variational Bayes with an Intractable Likelihood (VBIL). In that
approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
$\epsilon > 0$,   thus this method inherits the drawbacks of the ABC-likelihood
including the choice of summary statistics and the inefficiency in evaluating
the ABC likelihood for high-dimensional data and small $\epsilon$.

@@

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.
